Inferring probabilistic stellar rotation periods using Gaussian processes

### 2017-02-05 documents/ms.tex
first-pass tdm edits through section 4

Standard methods to measure rotation periods include detecting
peaks in a Lomb-Scargle \citep{Lomb1976, Scargle1982} (LS) periodogram
\citep[e.g.][]{Reinhold2013}, Auto-Correlation Functions (ACFs),
\citep{Mcquillan2013} and wavelet transforms \citep{Garcia2014}.
The precisions of the LS periodogram and wavelet methods are limited by the
suitability of the model choice: a sinusoid for the LS periodogram,
and a choice of mother wavelet, assumed to desribe the data
over a range of transpositions \citep[see, \eg][]{Carter2010},
for the wavelet method.
In contrast, the ACF method is much better suited to signals that are
non-sinusoidal.
In fact, as long as the signal is approximately periodic the ACF will
display a peak at the rotation period, no matter its shape.
A drawback of the ACF method, however, is that it requires data to be
evenly-spaced\footnote{\citet{Edelson1988} describe a method for computing
ACFs for unevenly-spaced data.}, which is not exactly the case with \Kepler\
light curves (although in many cases it can be approximated as uniformly
sampled) and not nearly the case for most ground-based measurements, such
as the future Large Synoptic Survey Telescope (\LSST).
An ACF is also an operation performed on the data rather than a generative
model of the data, and so is not inherently probabilistic.
This means that the effects of the observational uncertainties cannot be
formally propagated to constraints on the rotation period.
% Many rotation periods in the literature have been inferred by measuring the
% position of the first peak in an ACF, however this approach can be dangerous.
% The exponential decay in correlation can shift the peak position short-wards
% of its true value, leading to an underestimate of the period.
% We return to this point in section \textsection \ref{sec:discussion}.

@@

While using an uninformed prior on period (e.g., log-flat) often suffices,
we also develop a method to construct an informed period prior, based on the 
autocorrelation function (ACF). 
The ACF has proven to be very useful for measuring stellar rotation periods 
\citep{Mcquillan2012, Mcquillan2013, Mcquillan2014}; however, the
method has several shortcomings,
most notably the inability to deliver uncertainties, but also
the necessity of several heuristic choices,
such as a timescale on which to smooth the ACF,
how to define a peak, whether the first or second peak
gets selected, and what constitutes a secure detection.
While this paper presents a rotation period inference method
that avoids these shortcomings,
it seems prudent to still use information available from the ACF.
We thus use the ACF to define a prior on period,
which can help the posterior sampling converge on the true period.

@@

\subsection{Comparison with literature methods}
\label{sec:comparison}
\subsubsection{ACF}

@@

\subsubsection{LS periodogram}

@@

In order to test this rotation period inference method on real data,
we applied it to a set of \nkois\ \Kepler\ Object of Interest (KOI)
host stars, \nkoimcq\ of which had rotation periods previously measured by
the ACF method by \citet{Mcquillan2013}.  We use the pipeline-corrected flux
(\texttt{pdcsap\_flux} column in the \Kepler\ light curve table), median-normalized
and unit-subtracted, and mask out all known transiting planet candidate signals.
As with the simulated light curves, we randomly subsample each
light curve by a factor of 30 and split it into segments of about 300 points
for the purposes of evaluating the likelihood.  We also follow the same MCMC
fitting procedure as with the simulated data, using the ACF-based prior as before.

@@

\begin{table*}
\begin{center}
\caption{Priors and bounds on the natural logarithms of the GP model parameters,
        for \Kepler\ light curves}
\begin{tabular}{lcc}
Parameter & Prior & Bounds\\
    \hline
    $\ln A$ & $\mathcal N(-13, 5.7)$ & (-20, 0) \\
    $\ln l$ & $\mathcal N(5.0, 1.2)$ & (2, 8) \\
    $\ln \Gamma$ & $\mathcal N(1.9, 1.4)$ & (0, 3) \\
    $\ln \sigma$ & $\mathcal N(-17, 5)$ & (-20, 0) \\
    $\ln P $ & Uniform / ACF-based & ($\ln 0.5, \ln 100$) \\
\end{tabular}
\end{center}
\end{table*}
\label{tab:koipriors}

Figure \ref{fig:mcquillan} compares the periods inferred with the GP method to
the ACF-based periods from \citet{Mcquillan2013} for the \nkoimcq\ overlapping
KOIs.
This comparison shows generally very good agreement, with only a few
exceptions, demonstrating that this method works not only on simulated data,
but also on real data---with the caveat that for any particular data set, some
care is needed regarding the setting the priors and ranges for the GP
hyperparameters.

\begin{figure}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/comparison_koi_02_03.pdf}
\caption[Comparison with McQuillan results.]
{A comparison of our GP rotation period measurements to those of
\citet{Mcquillan2014}.
The data points are coloured by the range of variability measured by
    \citet{Mcquillan2014}, defined as the interval between the 5th and 95th
    percentiles of normalized flux per period bin in millimagnitudes.
    The results show good agreement.}
\label{fig:mcquillan}
\end{center}
\end{figure}

>>>>

Standard methods to measure rotation periods include detecting
peaks in a Lomb-Scargle \citep{Lomb1976, Scargle1982} (LS) periodogram
\citep[e.g.][]{Reinhold2013}, Auto-Correlation Functions (ACFs),
\citep{Mcquillan13b} and wavelet transforms \citep{Garcia2014}.
The precisions of the LS periodogram and wavelet methods are limited by the
suitability of the model choice: a sinusoid for the LS periodogram,
and a choice of mother wavelet, assumed to desribe the data
over a range of transpositions \citep[see, \eg][]{Carter2010},
for the wavelet method.
In contrast, the ACF method is much better suited to signals that are
non-sinusoidal.
In fact, as long as the signal is approximately periodic the ACF will
display a peak at the rotation period, no matter its shape.
A drawback of the ACF method, however, is that it requires data to be
evenly-spaced\footnote{\citet{Edelson1988} describe a method for computing
ACFs for unevenly-spaced data.}, which is not exactly the case with \Kepler\
light curves (although in many cases it can be approximated as uniformly
sampled) and not nearly the case for most ground-based measurements, such
as the future Large Synoptic Survey Telescope (\LSST).
An ACF is also an operation performed on the data rather than a generative
model of the data, and so is not inherently probabilistic.
This means that the effects of the observational uncertainties cannot be
formally propagated to constraints on the rotation period.
% Many rotation periods in the literature have been inferred by measuring the
% position of the first peak in an ACF, however this approach can be dangerous.
% The exponential decay in correlation can shift the peak position short-wards
% of its true value, leading to an underestimate of the period.
% We return to this point in section \textsection \ref{sec:discussion}.

@@

While using an uninformed prior on period (e.g., log-flat) often suffices,
we also develop a method to construct an informed period prior, based on the 
autocorrelation function (ACF). 
The ACF has proven to be very useful for measuring stellar rotation periods 
\citep{Mcquillan2012, Mcquillan13b, Mcquillan2014}; however, the
method has several shortcomings,
most notably the inability to deliver uncertainties, but also
the necessity of several heuristic choices,
such as a timescale on which to smooth the ACF,
how to define a peak, whether the first or second peak
gets selected, and what constitutes a secure detection.
While this paper presents a rotation period inference method
that avoids these shortcomings,
it seems prudent to still use information available from the ACF.
We thus use the ACF to define a prior on period,
which can help the posterior sampling converge on the true period.

@@

\subsection{Comparison with literature methods}
\label{sec:comparison}
\subsubsection{ACF}
\label{sec:acf}

@@

\subsubsection{LS periodogram}
\label{sec:ls}

@@

In order to test our rotation period inference method on real data,
we apply it to a set of \nkois\ \Kepler\ Object of Interest (KOI)
host stars, \nkoimcq\ of which had rotation periods previously measured by
the ACF method by \citet{Mcquillan2013}.  We use the pipeline-corrected flux
(\texttt{pdcsap\_flux} column in the \Kepler\ light curve table), median-normalized
and unit-subtracted, and mask out all known transiting planet candidate signals.
As with the simulated light curves, we randomly subsample each
light curve by a factor of 30 and split it into segments of about 300 points
for the purposes of evaluating the likelihood.  We also follow the same MCMC
fitting procedure as with the simulated data, using the ACF-based prior as before.

@@

\begin{table*}
\begin{center}
\caption{Priors and bounds on the natural logarithms of the GP model parameters,
        for \Kepler\ light curves}
\begin{tabular}{lcc}
Parameter & Prior & Bounds\\
    \hline
    $\ln A$ & $\mathcal N(-13, 5.7)$ & (-20, 0) \\
    $\ln l$ & $\mathcal N(5.0, 1.2)$ & (2, 8) \\
    $\ln \Gamma$ & $\mathcal N(1.9, 1.4)$ & (0, 3) \\
    $\ln \sigma$ & $\mathcal N(-17, 5)$ & (-20, 0) \\
    $\ln P $ & ACF-based & ($\ln 0.5, \ln 100$) \\
\end{tabular}
\end{center}
\end{table*}
\label{tab:koipriors}

Figure \ref{fig:mcquillan} compares the periods inferred with the GP method to
the ACF-based periods from \citet{Mcquillan2013} for the \nkoimcq\ overlapping
KOIs.
This comparison shows generally very good agreement, with only a few
exceptions, demonstrating that this method works not only on simulated data,
but also on real data---with the caveat that for any particular data set, some
care is needed regarding the setting the priors and ranges for the GP
hyperparameters.
Notably, below $\ln P$ of about 2.5, the GP method recovers periods 
systematically slightly larger than \citet{Mcquillan2013}---a likely consequence
of correcting for the ACF peak measurement bias discussed in Section \ref{sec:acf}.

\begin{figure}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/comparison_koi_02_03.pdf}
\caption[Comparison with McQuillan results.]
{A comparison of our GP rotation period measurements to those of
\citet{Mcquillan2013}.
The data points are coloured by the range of variability measured by
    \citet{Mcquillan2013}, defined as the interval between the 5th and 95th
    percentiles of normalized flux per period bin in millimagnitudes.
    The results show good agreement.}
\label{fig:mcquillan}
\end{center}
\end{figure}
